---
title: "Final Project"
author: "Jördis Strack, Julia King"
format: html
editor: source
---

## 0. Setup Chunk

Text

```{r}
#| output: false
if (!require("pacman")) install.packages("pacman")
pacman::p_load(readr, tidyverse, class, caret, bench, DataExplorer, randomForest)
```

## 1. Data Exploration

Text

```{r}
#| warning: false
# import dataset
data = read_csv("data/BreastCancer.csv", show_col_types = FALSE)
```

```{r}
# first overview
head(data)
```


```{r}
# explore features and labels
str(data)
```


```{r}
# continue exploration
dim(data)
```

```{r}
# count missing values per column
missing_count_per_column <- colSums(is.na(data))
missing_count_per_column

# INTERPRETATION: There are only missing values for variable
# data$...33 -> caused by trailing comma in csv

data <- subset(data, select = -c(...33))
```

```{r}
#| fig-width: 9
#| fig-height: 7
# visualize distribution of classes

# visualize means
data_means = pivot_longer(data %>% select(3:12), 
                                  everything(), 
                                  names_to = "Variable", 
                                  values_to = "Value")
ggplot(data_means, aes(x = Value)) +
  geom_density(fill = "skyblue", color = "black", alpha = 0.5) +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Density Plots of the variable means")
```

Most means seem to be right-skewed. Thus, they will likely require some form of correction. Taking the logarithm is infeasible because of 0 values. Thus, square roots are shown below:

```{r}
#| fig-width: 9
#| fig-height: 7
ggplot(data_means, aes(x = sqrt(Value))) +
  geom_density(fill = "red", color = "black", alpha = 0.5) +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Density Plots of the square root transformation of the variable means")
```


```{r}
#| fig-width: 9
#| fig-height: 7
# visualize standard errors
data_variances = pivot_longer(data %>% select(13:22), 
                                  everything(), 
                                  names_to = "Variable", 
                                  values_to = "Value")
ggplot(data_variances, aes(x = Value)) +
  geom_density(fill = "skyblue", color = "black") +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Density Plots of the variable variances")
```
Similarly, the standard errors are also right-skewed. They could be fixed with a log-transformation, as standard errors are by definition > 0 (assuming there is variation).

```{r}
#| fig-width: 9
#| fig-height: 7
ggplot(data_variances, aes(x = log(Value))) +
  geom_density(fill = "red", color = "black", alpha = 0.5) +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Density Plots of the log transformations of the variable variances")
```

```{r}
#| fig-width: 9
#| fig-height: 7
# visualize highest values
data_worst = pivot_longer(data %>% select(23:32), 
                                  everything(), 
                                  names_to = "Variable", 
                                  values_to = "Value")
ggplot(data_worst, aes(x = Value)) +
  geom_density(fill = "skyblue", color = "black") +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Density Plots of the highest values of the variables")
```

```{r}
#| fig-width: 9
#| fig-height: 7
ggplot(data_worst, aes(x = sqrt(Value))) +
  geom_density(fill = "red", color = "black", alpha = 0.5) +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Density Plots of the square root transformation of the highest values of the variables")
```

```{r}
#| fig-width: 9
#| fig-height: 7
plot_intro(data)
plot_correlation(data)
```


```{r}
# compute descriptive statistics
summary(data)

# compute share of malignant cancers
malignant_share = sum(data$diagnosis == "M") / nrow(data)
print(paste("The share of malignant cancers in the dataset is ", round(malignant_share, 3), ".", sep = ""))
```

## 2. Data Preprocessing

Text

```{r}
# handle categorical variables appropriately, if present

# change diagnosis to binary variable for easier interpretation
data = data %>% 
  mutate(malignant = as.factor(if_else(diagnosis == "M", TRUE, FALSE))) %>%
  select(-diagnosis) %>%
  relocate(malignant, .before = 2)
```

```{r}
# perform variable selection, if required

# remove id column
data = data %>% select(-id)

# rename variables with whitespaces
colnames(data) = sub(" ", "_", colnames(data))

data = as.data.frame(data, names = TRUE)
```

```{r}
# correlation analysis

# calculate correlation matrix
# correlation_matrix <- cor(data[,2:31])
# correlation_matrix

# find attributes that are highly corrected (ideally >0.75)
# highly_correlated <- findCorrelation(correlation_matrix, cutoff=0.5)
# highly_correlated
```

As seen above in step 1.4) we should most likely reduce the number of variables used for our classifier. This fits the results presented by the heat map of all variabels

```{r}
#| fig-width: 9
#| fig-height: 7
# prepare training scheme
control <- trainControl(method="repeatedcv", number=10, repeats=3)

# train the model
model <- train(malignant~., data=data, method="lvq", preProcess="scale", 
                            trControl=control)

# estimate variable importance
importance <- varImp(model, scale=FALSE)

# summarize importance
importance

# plot importance
plot(importance)
```

```{r}
# standardize features
standardize_col <- function(col) {
    return(col - mean(col)) / sd(col)
  }

data = data %>%
  mutate_at(vars(-malignant), ~ standardize_col(.))
```


```{r}
# create training and testing sets
set.seed(-12)
train_size = floor(nrow(data) * 0.8)
train_index = sample(1:nrow(data), train_size, replace = FALSE)
test_index = setdiff(1:nrow(data), train_index)

# prepare as dataframes
train_data = data[train_index, ]
test_data = data[test_index, ]

# prepare matrixes and vectors
train_data_x = train_data %>% select(-malignant) %>% as.matrix()
train_data_y = train_data %>% mutate(malignant = as.logical(malignant)) %>% pull(malignant)
test_data_x = test_data %>% select(-malignant) %>% as.matrix()
test_data_y = test_data %>% mutate(malignant = as.logical(malignant)) %>% pull(malignant)

print(paste("The training set contains", nrow(train_data), "entries."))
print(paste("The testing set contains", nrow(test_data), "entries."))
```


## 3. Model Selection & Model Evaluation

### 3.1 K-Nearest Neighbors

Text

# define evaluation method for KNN for given dataset
```{r}
evaluate_KNN <- function(train_x, train_y, test_x, test_y, k_min = 1, k_max) {
  k_range = k_min:k_max
  mse = c()
  accuracy = c()
  precision = c()
  recall = c()
  n = nrow(test_x)
  for (k in k_range) {
    test_y_hat = knn(train = train_x, 
                     test = test_x, 
                     cl = train_y, 
                     k = k)
    test_y_hat = as.logical(test_y_hat)
    pred_matrix = table(Actual = test_y, Predicted = test_y_hat)
    # columns = predicted, rows = actual, 1 for false, 2 for true. EX: pred_matrix[1, 2] returns false positives
    mse = c(mse, mean((test_y - test_y_hat)^2))
    accuracy = c(accuracy, (pred_matrix[1, 1] + pred_matrix[2, 2]) / n)
    precision = c(precision, (pred_matrix[2, 2]) / (pred_matrix[2, 2] + pred_matrix[1, 2]))
    recall = c(recall, (pred_matrix[2, 2]) / (pred_matrix[2, 2] + pred_matrix[2, 1]))
  }
  results = data.frame(k_range, mse, accuracy, precision, recall)
  return(results)
}
evaluate_KNN(train_x = train_data_x, train_y = train_data_y, test_x = test_data_x, test_y = test_data_y, k_max = 11)
```



```{r}

# choose model parameters (K) using cross validation
set.seed(-12)

n = nrow(train_data_x)
K_max = floor(nrow(train_data_x) * 0.25) # performance will most likely not improve by defining K above 25% of the dataset 
nfolds = 10  # chosen to minimize variance while keeping bias low
permutation = sample(1:n)  
MSE_fold = matrix(0, nfolds, K_max)  
accuracy_fold = matrix(0, nfolds, K_max) 
recall_fold = matrix(0, nfolds, K_max) 
precision_fold = matrix(0, nfolds, K_max) 

# getting percentage correct and 
for (j in 1:nfolds) {
    test_index = permutation[floor((j-1)*n/nfolds+1) : floor(j*n/nfolds)]  
    train_index = setdiff(1:n, test_index) 
    for (K in 1:K_max) {
        y_hat = knn(train = train_data_x[train_index, ], 
                    test = train_data_x[test_index, ], 
                    cl = train_data_y[train_index], 
                    k = K)
        y_hat = as.logical(y_hat)
        MSE_fold[j,K] = mean((train_data_y[test_index] - y_hat)^2)  
        accuracy_fold[j,K] = mean((train_data_y[test_index] == y_hat))
        recall_fold[j, K] = sum((train_data_y[test_index] == y_hat) & (y_hat == TRUE)) / sum((train_data_y[test_index] == TRUE))
        precision_fold[j, K] = sum((train_data_y[test_index] == y_hat) & (y_hat == TRUE)) / sum((y_hat == TRUE))
    }
}
MSE_cv = colMeans(MSE_fold)
accuracy = colMeans(accuracy_fold)
recall = colMeans(recall_fold)
precision = colMeans(precision_fold)
knn_evaluation = tibble(k = 1:K_max, accuracy, precision, recall)
```

```{r}
#| fig-width: 9
#| fig-height: 7
# visualize results for different K:
ggplot(knn_evaluation, aes(x = k)) +
  geom_line(aes(y = accuracy), 
            color = "blue") +
  geom_line(aes(y = precision), 
            color = "red") +
  geom_line(aes(y = recall), 
            color = "green") + 
  geom_point(mapping = aes(x = which.max(accuracy), y = max(accuracy)), 
             color = "blue", size = 3) + 
  geom_point(mapping = aes(x = which.max(precision), y = max(precision)), 
             color = "red", size = 3) + 
  geom_point(mapping = aes(x = which.max(recall), y = max(recall)), 
             color = "green", size = 3) + 
  geom_text(mapping = aes(x = which.max(accuracy), y = max(accuracy), 
            label = which.max(accuracy)), 
            vjust = 1, hjust = 0) + 
  geom_text(mapping = aes(x = which.max(precision), y = max(precision), 
            label = which.max(precision)), 
            vjust = 1, hjust = 0) + 
  geom_text(mapping = aes(x = which.max(recall), y = max(recall), 
            label = which.max(recall)), 
            vjust = 1, hjust = 0)
```

### 3.2 Random Forest

```{r}
# set up model on training data
RF_model <- randomForest(
  formula = malignant ~ .,
  data = train_data
)

RF_model
```

Looks pretty good! Train data set model accuracy is roughly 96,26%. Go ahead to prediction and confusion matrix

```{r}
pred_RF_train <- predict(RF_model, train_data)
confusionMatrix(pred_RF_train, train_data$malignant)
```

Mini analysis:

```{r}
#| fig-width: 9
#| fig-height: 7
# visualize the tree size 
hist(treesize(RF_model),
     main = "Nodes used for tree construction",
     col = "palegreen")

# again, look at importance of predictors
varImpPlot(RF_model,
           sort = T,
           n.var = 30,
           main = "Importance of predictors")

importance(RF_model)
```

Wow, this looks so different!!

```{r}
# prediction for test data
# pred_RF_test <- predict(RF_model, test_data)
# confusionMatrix(pred_RF_test, test_data$malignant)
```

Setup of cross-validation

```{r}
nfolds = 10

accuracy_fold <- numeric(nfolds)
precision_fold <- numeric(nfolds)
recall_fold <- numeric(nfolds)
f1_fold <- numeric(nfolds)

# Prepare train/test splits for Cross Validation
set.seed(-12)
permutation = sample(1:nrow(train_data))

# perform Cross Validation

for (j in 1:nfolds){
  test_indices <- permutation[((j - 1) * nrow(train_data) / nfolds + 1) : (j * nrow(train_data) / nfolds)]
  train_indices <- setdiff(1:nrow(train_data), test_indices)

  # obtain training and testing folds
  train_fold <- train_data[train_indices, ]
  test_fold <- train_data[test_indices, ]
  
  # fit Random Forest model on training fold
  RF_model <- randomForest(
    formula = malignant ~ .,
    data = train_fold
  )
  
  # predict on pseudo-test folds
  pred <- predict(RF_model, test_fold)
  
  # obtain evaluation metrics for the fold
  cm <- confusionMatrix(data = pred, reference = test_fold$malignant)
  accuracy_fold[j] <- cm$overall["Accuracy"]
  precision_fold[j] <- cm$byClass["Precision"][1]  # precision for class 'M'
  recall_fold[j] <- cm$byClass["Recall"][1]  # recall for class 'M'
  f1_fold[j] <- cm$byClass["F1"][1]  # F1 Score for class 'M'
}

# compute mean of evaluation metrics across folds
mean_accuracy <- mean(accuracy_fold)
mean_precision <- mean(precision_fold)
mean_recall <- mean(recall_fold)
mean_f1 <- mean(f1_fold)

cat("Mean Accuracy:", mean_accuracy, "\n")
cat("Mean Precision:", mean_precision, "\n")
cat("Mean Recall:", mean_recall, "\n")
cat("Mean F1 Score:", mean_f1, "\n")
```

## 4. Fine-Tuning

Text

```{r}
tuned_RF_model <- tuneRF(train_data[,-5], train_data[,5],
            stepFactor = 0.5,
            plot = TRUE,
            ntreeTry = 110, # <- I've played around with it & 110 seems to be highest
            trace = TRUE,
            improve = 0.05)
```

## 5. Final Model Selection

Text

```{r}

```