---
title: "Final Project"
author: "Jördis Strack, Julia King"
format: html
editor: source
---

## 0. Setup Chunk

<<<<<<< HEAD
Text

```{r}
#| output: false
if (!require("pacman")) install.packages("pacman")
pacman::p_load(readr, tidyverse, class)
```

=======
```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(readr)
```


>>>>>>> 9c4c2b9d73b9e7c1d1a033cf9a9614dcadb0439a
## 1. Data Exploration

Text

```{r}
#| warning: false
# import dataset
breastcancer = read_csv("data/BreastCancer.csv", show_col_types = FALSE) %>%
  select(-last_col())
# remove irrelevant column added by trailing commas in file
```

```{r}
# explore features and labels
str(breastcancer)
```

```{r}
# continue exploration
names(breastcancer)
summary(breastcancer)
```



```{r}
# check for missing values
na_sum = sum(is.na(breastcancer))
print(paste(na_sum, "NA's found."))
```

```{r}
#| fig-width: 9
#| fig-height: 7
# visualize distribution of classes

# visualize means
breastcancer_means = pivot_longer(breastcancer %>% select(3:12), 
                                  everything(), 
                                  names_to = "Variable", 
                                  values_to = "Value")
ggplot(breastcancer_means, aes(x = Value)) +
  geom_density(fill = "skyblue", color = "black", alpha = 0.5) +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Density Plots of the variable means")
```

Most means seem to be right-skewed. Thus, they will likely require some form of correction. Taking the logarithm is infeasible because of 0 values. Thus, square roots are shown below:

```{r}
#| fig-width: 9
#| fig-height: 7
ggplot(breastcancer_means, aes(x = sqrt(Value))) +
  geom_density(fill = "red", color = "black", alpha = 0.5) +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Density Plots of the square root transformation of the variable means")
```


```{r}
#| fig-width: 9
#| fig-height: 7
# visualize standard errors
breastcancer_variances = pivot_longer(breastcancer %>% select(13:22), 
                                  everything(), 
                                  names_to = "Variable", 
                                  values_to = "Value")
ggplot(breastcancer_variances, aes(x = Value)) +
  geom_density(fill = "skyblue", color = "black") +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Density Plots of the variable variances")
```
Similarly, the standard errors are also right-skewed. They could be fixed with a log-transformation, as standard errors are by definition > 0 (assuming there is variation).

```{r}
#| fig-width: 9
#| fig-height: 7
ggplot(breastcancer_variances, aes(x = log(Value))) +
  geom_density(fill = "red", color = "black", alpha = 0.5) +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Density Plots of the log transformations of the variable variances")
```

```{r}
#| fig-width: 9
#| fig-height: 7
# visualize highest values
breastcancer_worst = pivot_longer(breastcancer %>% select(23:32), 
                                  everything(), 
                                  names_to = "Variable", 
                                  values_to = "Value")
ggplot(breastcancer_worst, aes(x = Value)) +
  geom_density(fill = "skyblue", color = "black") +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Density Plots of the highest values of the variables")
```

```{r}
#| fig-width: 9
#| fig-height: 7
ggplot(breastcancer_worst, aes(x = sqrt(Value))) +
  geom_density(fill = "red", color = "black", alpha = 0.5) +
  facet_wrap(~Variable, scales = "free") +
  labs(title = "Density Plots of the square root transformation of the highest values of the variables")
```


```{r}
# compute descriptive statistics
summary(breastcancer)
malignant_share = sum(breastcancer$diagnosis == "M") / nrow(breastcancer)
print(paste("The share of malignant cancers in the dataset is ", round(malignant_share, 3), ".", sep = ""))
```


## 2. Data Preprocessing

Text

```{r}
# handle categorical variables appropriately, if present

# change diagnosis to binary variable for easier interpretation
breastcancer = breastcancer %>% 
  mutate(malignant = if_else(diagnosis == "M", TRUE, FALSE)) %>%
  select(-diagnosis) %>%
  relocate(malignant, .before = 2)
```

```{r}
# perform variable selection, if required

# remove id column
breastcancer = breastcancer %>% select(-id)

# other variables will be kept for now as they could theoretically all contribute to the result.
```

```{r}
# standardize features
standardize_col <- function(col) {
    return(col - mean(col)) / sd(col)
  }

breastcancer = breastcancer %>%
  mutate_at(vars(-malignant), ~ standardize_col(.))
```


```{r}
# create training and testing sets
set.seed(-12)
train_size = floor(nrow(breastcancer) * 0.8)
train_index = sample(1:nrow(breastcancer), train_size, replace = FALSE)
test_index = setdiff(1:nrow(breastcancer), train_index)

# prepare as dataframes
bc_train = breastcancer[train_index, ]
bc_test = breastcancer[test_index, ]

# prepare matrixes and vectors
bc_train_x = bc_train %>% select(-malignant) %>% as.matrix()
bc_train_y = bc_train %>% pull(malignant)
bc_test_x = bc_test %>% select(-malignant) %>% as.matrix()
bc_test_y = bc_test %>% pull(malignant)

print(paste("The training set contains", nrow(bc_train), "entries."))
print(paste("The testing set contains", nrow(bc_test), "entries."))
```


## 3. Model Selection & Model Evaluation

### 3.1 K-Nearest Neighbors

Text

```{r}
# choose model parameters (K) using cross validation
set.seed(-12)

n = nrow(bc_train_x)
K_max = floor(nrow(bc_train_x) * 0.25) # performance will most likely not improve by defining K above 25% of the dataset 
nfolds = 10  # chosen to minimize variance while keeping bias low
permutation = sample(1:n)  
MSE_fold = matrix(0, nfolds, K_max)  
accuracy_fold = matrix(0, nfolds, K_max) 
recall_fold = matrix(0, nfolds, K_max) 
precision_fold = matrix(0, nfolds, K_max) 

# getting percentage correct and 
for (j in 1:nfolds) {
    test_index = permutation[floor((j-1)*n/nfolds+1) : floor(j*n/nfolds)]  
    train_index = setdiff(1:n, test_index) 
    for (K in 1:K_max) {
        y_hat = knn(train = bc_train_x[train_index, ], 
                    test = bc_train_x[test_index, ], 
                    cl = bc_train_y[train_index], 
                    k = K)
        y_hat = as.logical(y_hat)
        MSE_fold[j,K] = mean((bc_train_y[test_index] - y_hat)^2)  
        accuracy_fold[j,K] = mean((bc_train_y[test_index] == y_hat))
        recall_fold[j, K] = sum((bc_train_y[test_index] == y_hat) & (y_hat == TRUE)) / sum((bc_train_y[test_index] == TRUE))
        precision_fold[j, K] = sum((bc_train_y[test_index] == y_hat) & (y_hat == TRUE)) / sum((y_hat == TRUE))
    }
}
MSE_cv = colMeans(MSE_fold)
accuracy = colMeans(accuracy_fold)
recall = colMeans(recall_fold)
precision = colMeans(precision_fold)
evaluation = tibble(k = 1:K_max, accuracy, precision, recall)
```

```{r}
# visualize results for different K:
ggplot(evaluation, aes(x = k)) +
  geom_line(aes(y = accuracy), color = "blue") +
  geom_line(aes(y = precision), color = "red") +
  geom_line(aes(y = recall), color = "green") + 
  geom_point(mapping = aes(x = which.max(accuracy), y = max(accuracy)), color = "blue", size = 3) + 
  geom_point(mapping = aes(x = which.max(precision), y = max(precision)), color = "red", size = 3) + 
  geom_point(mapping = aes(x = which.max(recall), y = max(recall)), color = "green", size = 3) + 
  geom_text(mapping = aes(x = which.max(accuracy), y = max(accuracy), label = which.max(accuracy)), vjust = 1, hjust = 0) + 
  geom_text(mapping = aes(x = which.max(precision), y = max(precision), label = which.max(precision)), vjust = 1, hjust = 0) + 
  geom_text(mapping = aes(x = which.max(recall), y = max(recall), label = which.max(recall)), vjust = 1, hjust = 0)
```


### 3.2 Random Forest

Text

```{r}

```

## 4. Fine-Tuning

Text

```{r}

```

## 5. Final Model Selection

Text

```{r}

```